Computational methods in landscape ecology

### 2024-10-18 Sections/01-introduction.qmd

Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land-use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land-use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modeling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].

### 2024-10-18 Sections/03-spatial-patterns.qmd

Traditionally and at present, many prominent approaches that quantify spatial patterns revolve around raster data using categorical values based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
<!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
The strength of landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
However, also limitations have been identified. 
These limitations include shortcomings in their ability to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

@@

The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape, omitting the spatial configuration.
Shannon's entropy can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and interclass distances [@Claramunt2012SpatiotemporalForm].
However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
@Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix which could be used to calculate various entropy measures.
This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).

Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, in the last few years, there has been a surge in the development of entropy-based metrics for landscape ecology.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
<!-- R1: the edge length -> Could you specify a bit -->
@Cushman2016CalculatingConfigurational proposed to relate the edge length (defined as the side lengths of neighboring cells with different LULC classes) to the microstate of the landscape and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
The relationship between Shannon and Boltzmann entropies in landscape ecology remains contentious, with recent studies challenging the thermodynamic interpretation of Boltzmann-inspired measures and advocating for Shannon entropy as a more general form [@Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial], highlighting the need for further research to elucidate their connections to environmental processes.

@@

As hundreds of metrics have already been developed for raster data, the most straightforward approach may be to reimplement these same metrics for vector data.
Although this is possible for many metrics, there are also metrics specifically related to the corresponding data models [@Yao2022VecLIFramework].
For example, in urban planning vector-based metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
Earlier approaches to quantifying shape complexity included four categories of compactness measures: perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].
However, also vector-based metrics have limitations.
The most important limitation relates to computational complexity, which makes calculations of vector-based metrics slower than their raster equivalents.
Another technical issue is the requirement for topologically correct data (e.g., geometries cannot overlap), which is often problematic, especially for data from different sources.
Last, the pattern-process link for vector-based metrics remains underexplored compared to raster-based metrics.

>>>>

Traditionally and at present, many prominent approaches that quantify spatial patterns revolve around raster data using categorical values based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
<!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
The strength of landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing derived products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
However, also limitations have been identified for many landscape metrics. 
These limitations include shortcomings to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

@@

The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape, omitting the spatial configuration.
Shannon's entropy can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and interclass distances [@Claramunt2012SpatiotemporalForm].
However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
@Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix which could be used to calculate various entropy measures.
This includes conditional entropy (representing complexity in configuration), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).

Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, in the last few years, there has been a surge in the development of entropy-based metrics for landscape ecology.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of 'macrostate' (the general state of a system) and 'microstate' (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
<!-- R1: the edge length -> Could you specify a bit -->
@Cushman2016CalculatingConfigurational proposed to relate the edge length (defined as the side lengths of neighboring cells with different LULC classes) to the microstate of the landscape and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
The relationship between Shannon and Boltzmann entropies in landscape ecology remains contentious, with recent studies challenging the thermodynamic interpretation of Boltzmann-inspired measures and advocating for Shannon entropy as a more general form [@Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial], highlighting the need for further research to elucidate their connections to environmental processes.

@@

As hundreds of metrics have already been developed for gridded raster data, the most straightforward approach may be to reimplement these same metrics for vector data. Although this is possible for many metrics, there are also metrics specifically related to the corresponding data models [@Yao2022VecLIFramework].
For example, in urban planning vector-based metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
Earlier approaches to quantifying shape complexity included four categories of compactness measures: perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].
However, also vector-based metrics have limitations.
The most important limitation relates to computational complexity, which makes calculations of vector-based metrics slower than their raster equivalents.
Another technical issue is the requirement for topologically correct data (e.g., geometries cannot overlap), which is often problematic, especially for data from different sources.
Last, the pattern-process link for vector-based metrics remains underexplored compared to raster-based metrics.

### 2024-10-18 Sections/08-spatial-planning.qmd

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
In the context of landscape ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
For example, there have been several recent advances in considering landscape patterns directly in spatial planning or through using simulation model outputs as input features [@Daigle2020OperationalizingEcological].

@@

A second development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
<!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found. Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
However, this comes with the drawbacks of reduced interpretability and because it is not an exact approach no guarantee that the best possible solution can been found.

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
<!-- R1: Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one. Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others. Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances. They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers. A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem. There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
Unfortunately, there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing] and the optimal solver will depend on the specific problem. 
However, due to software licenses and related cost the optimal solver might not be always accessible.

>>>>

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState;@hanson2024systematic].
In the context of landscape ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
For example, there have been several recent advances in considering landscape patterns directly in spatial planning or through using simulation model outputs as input features [@Daigle2020OperationalizingEcological].

@@

A second development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
<!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found. Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
<!-- MJ: Broadly agree (although similar can be said for many other heuristical approaches). The main benefit is in the use of dynamic and non-linear functions which would not be computationally feasible with exact algorithms. -->
However, this comes with the drawbacks of reduced interpretability, and similar to traditional spatial planning approaches such as Marxan, it relies on heuristics that can not guarantee that a best possible solution can been found [@Hanson2019OptimalityPrioritizing].

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning. Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
<!-- R1: Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one. Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others. Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances. They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers. A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem. There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
<!-- MJ: Agree to some extent, although again there are many comparisons that have shown that commercial solvers will usually outcompete open-source solvers.--->
Unfortunately, there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing] and the performance of any individual solvers can depend on the specific problem, data and computational resources. Nevertheless, existing comparisons often show that proprietary solvers can be used to obtain solutions to complex problems in reasonable time periods [@hanson2024systematic], however, due to software licenses and related cost certain solvers might not always be openly accessible.
